---
title: "Statistical analysis of robot behavior and perception data"
author: "Maria Teresa Parreira, Sarah Gillet, Katie Winkel, Iolanda Leite"
date: "2023-01-14"
output:
  pdf_document: default
  html_document: default
---

R script for 

*M. T. Parreira, S. Gillet, K. Winkle and I. Leite, "How Did We Miss This? A Case Study on Unintended Biases in Robot Social Behavior," 2023 18th ACM/IEEE International Conference on Human-Robot Interaction (HRI), Stockholm, Sweden, 2023*

Data sources in [this repository](https://github.com/mteresaparreira/gender-biased-robot-duck). 
```{r}
## Imports


if (!require("car")) install.packages("car") 
if (!require("pastecs")) install.packages("pastecs")

library(tidyverse)
library(lme4)
library(car)
library(pastecs)
```



# General Analyses

## Perception of Robot

```{r}
questionnaires_org <- read_csv("df_postq.csv")

#filter for 2 cond
questionnaires_org <- questionnaires_org %>% filter(condition != "rubberduck")

questionnaires <- questionnaires_org %>% 
  mutate(condition = factor(condition, levels=c('naive','datadriven'))) %>%
  mutate(listening_behavior = as.numeric(listening_behavior))%>%
  mutate(closeness = as.numeric(closeness))%>%
  mutate(competence = as.numeric(competence))%>%
  mutate(warmth = as.numeric(warmth))%>%
  mutate(discomfort = as.numeric(discomfort))

#checking data normality and homogeneity
by(questionnaires[, c("listening_behavior","closeness","competence","warmth","discomfort")], questionnaires$condition, stat.desc, basic = FALSE,norm = TRUE)

print("listening_behavior")
leveneTest(questionnaires$listening_behavior,questionnaires$condition)
print("closeness")
leveneTest(questionnaires$closeness,questionnaires$condition)
print("competence")
leveneTest(questionnaires$competence,questionnaires$condition)
print("warmth")
leveneTest(questionnaires$warmth,questionnaires$condition)
print("discomfort")
leveneTest(questionnaires$discomfort,questionnaires$condition)

#perception of listener behavior (from Murray et al. (2021))
model1 <-aov(listening_behavior ~ condition, data = questionnaires)
Anova(model1, type="III")
#plot(model1)

#perception of closeness with robot (from Murray et al. (2021))
model1 <-aov(closeness ~ condition, data = questionnaires)
Anova(model1, type="III")
#plot(model1)

#RoSAS 
model1 <-aov(competence ~ condition, data = questionnaires)
Anova(model1, type="III")
#plot(model1)

#RoSAS (not normal)
model1<-wilcox.test(warmth ~ condition, data = questionnaires, exact = FALSE,
correct= FALSE)
model1

#RoSAS (not normal)
model1<-wilcox.test(discomfort ~ condition, data = questionnaires, exact = FALSE,
correct= FALSE)
model1
```

# Controlling for Effect of Gender

## Robot Listening Behavior and Robot Perception Analyses
```{r}
questionnaires_org <- read_csv('final_robot_logs.csv')


#filter for 2 cond
questionnaires_orgn <- questionnaires_org %>% filter(condition == "naive")
questionnaires_orgd<- questionnaires_org %>% filter(condition == "datadriven")

questionnairesn <- questionnaires_orgn %>% 
  mutate(gender = factor(gender)) %>%
  mutate(talk_ratio = as.numeric(talk_ratio))%>%
  mutate(listening_behavior = as.numeric(listening_behavior))%>%
  mutate(closeness = as.numeric(closeness))%>%
  mutate(competence = as.numeric(competence))%>%
  mutate(warmth = as.numeric(warmth))%>%
  mutate(discomfort = as.numeric(discomfort))%>%
  mutate(bc_freq = as.numeric(bc_freq_all))
  
questionnairesd <- questionnaires_orgd %>% 
  mutate(gender = factor(gender)) %>%
  mutate(talk_ratio = as.numeric(talk_ratio))%>%
  mutate(listening_behavior = as.numeric(listening_behavior))%>%
  mutate(closeness = as.numeric(closeness))%>%
  mutate(competence = as.numeric(competence))%>%
  mutate(warmth = as.numeric(warmth))%>%
  mutate(discomfort = as.numeric(discomfort))%>%
  mutate(bc_freq = as.numeric(bc_freq_all))

#checking stats per gender (NaiveL condition)
by(questionnairesn[, c("bc_freq", "talk_ratio","listening_behavior","closeness","competence","warmth","discomfort")], questionnairesn$gender, stat.desc, basic = FALSE,norm = TRUE)

print("listening_behavior")
leveneTest(questionnairesn$listening_behavior,questionnairesn$gender)
print("closeness")
leveneTest(questionnairesn$closeness,questionnairesn$gender)
print("competence")
leveneTest(questionnairesn$competence,questionnairesn$gender)
print("warmth")
leveneTest(questionnairesn$warmth,questionnairesn$gender)
print("discomfort")
leveneTest(questionnairesn$discomfort,questionnairesn$gender)
print("bc_freq")
leveneTest(questionnairesn$bc_freq,questionnairesn$gender)
print("talk_ratio")
leveneTest(questionnairesn$bc_freq,questionnairesn$gender)

#checking stats per gender (DataL condition)

by(questionnairesd[, c("bc_freq", "talk_ratio","listening_behavior","closeness","competence","warmth","discomfort")], questionnairesd$gender, stat.desc, basic = FALSE,norm = TRUE)

print("listening_behavior")
leveneTest(questionnairesd$listening_behavior,questionnairesd$gender)
print("closeness")
leveneTest(questionnairesd$closeness,questionnairesd$gender)
print("competence")
leveneTest(questionnairesd$competence,questionnairesd$gender)
print("warmth")
leveneTest(questionnairesd$warmth,questionnairesd$gender)
print("discomfort")
leveneTest(questionnairesd$discomfort,questionnairesd$gender)
print("bc_freq")
leveneTest(questionnairesd$bc_freq,questionnairesd$gender)
print("talk_ratio")
leveneTest(questionnairesd$bc_freq,questionnairesd$gender)

#bc frequency (NaiveL)
model1 <-aov(bc_freq ~ gender + talk_ratio, data = questionnairesn)
Anova(model1, type="III")
#plot(model1)

#bc frequency (DataL)
model1 <-aov(bc_freq ~ gender + talk_ratio, data = questionnairesd)
Anova(model1, type="III")

#ROBOT PERCEPTION (NaiveL) - listening behavior (from Murray et al. (2021))
model1 <-aov(listening_behavior ~ gender + bc_freq, data = questionnairesn)
Anova(model1, type="III")
#plot(model1)

#ROBOT PERCEPTION (DataL) - listening behavior (from Murray et al. (2021))
model1 <-aov(listening_behavior ~ gender + bc_freq, data = questionnairesd)
Anova(model1, type="III")
#plot(model1)

#ROBOT PERCEPTION (NaiveL) - closeness (from Murray et al. (2021))
model1 <-aov(closeness ~ gender + bc_freq, data = questionnairesn)
Anova(model1, type="III")
#plot(model1)

#ROBOT PERCEPTION (DataL) - closeness (from Murray et al. (2021))
model1 <-aov(closeness ~ gender + bc_freq, data = questionnairesd)
Anova(model1, type="III")
#plot(model1)

#ROBOT PERCEPTION (NaiveL) - Competence (from RoSAS)
model1 <-aov(competence ~ gender + bc_freq, data = questionnairesn)
Anova(model1, type="III")
#plot(model1)

#ROBOT PERCEPTION (DataL) - Competence (from RoSAS)
model1 <-aov(competence ~ gender + bc_freq, data = questionnairesd)
Anova(model1, type="III")
#plot(model1)

#ROBOT PERCEPTION (NaiveL) - Warmth (from RoSAS)
model1 <-aov(warmth ~ gender + bc_freq, data = questionnairesn)
Anova(model1, type="III")
#plot(model1)

#ROBOT PERCEPTION (DataL) - Warmth (from RoSAS)
model1 <-aov(warmth ~ gender + bc_freq, data = questionnairesd)
Anova(model1, type="III")
#plot(model1)

#ROBOT PERCEPTION (NaiveL) - Discomfort (from RoSAS)
model1 <-aov(discomfort ~ gender + bc_freq, data = questionnairesn)
Anova(model1, type="III")
#plot(model1)

#ROBOT PERCEPTION (DataL) - Discomfort (from RoSAS)
model1 <-aov(discomfort ~ gender + bc_freq, data = questionnairesd)
Anova(model1, type="III")
#plot(model1)
```

## NaiveL - Analysis of Pitch Distribution
```{r}
questionnaires_org <- read_csv("pitchesnewdf.csv")

questionnaires <- questionnaires_org %>% 
  mutate(gender = factor(gender))%>%
  mutate(rel_below_pitch = as.numeric(rel_below_pitch)) %>%
  mutate(below_avg_size = as.numeric(below_avg_size))

by(questionnaires[, c("rel_below_pitch","below_avg_size")], questionnaires$gender, stat.desc, basic = FALSE,norm = TRUE)

print("rel_below_pitch")
leveneTest(questionnaires$rel_below_pitch,questionnaires$gender)
print("below_avgsize")
leveneTest(questionnaires$below_avg_size,questionnaires$gender)

#samples below 26th percentile (LowPitch)
model1<-wilcox.test(rel_below_pitch ~ gender, data = questionnaires, exact = FALSE,
correct= FALSE)
model1

#consecutive duration below 26th percentile (LowPitchDuration)
model1<-wilcox.test(below_avg_size ~ gender, data = questionnaires, exact = FALSE,
correct= FALSE)
model1
```

## DataL - Principal Component Analysis of State Space
```{r}
questionnaires_org <- read_csv("pitchescompsdf.csv")

questionnaires <- questionnaires_org %>% 
  mutate(gender = factor(gender))%>%
  mutate(pc0 = as.numeric(PC0)) %>%
  mutate(pc1 = as.numeric(PC1))

#PC0
model1 <-aov(pc0 ~ gender, data = questionnaires)
Anova(model1, type="III")
#plot(model1)

#PC1
model1 <-aov(pc1 ~ gender, data = questionnaires)
Anova(model1, type="III")
#plot(model1)

```